A simulation model of DNA methylation data

### 2015-05-01 Rmarkdown/methsim.Rmd
Continue methsim chapter

\begin{chapabstract}
I present the `methsim` software to simulate DNA methylation data. `methsim` incorporates a model of within-fragment co-methylation and simulates individual sequencing reads, which sets it apart from existing software. I describe the statistical model behind `methsim` and its software implementation. Finally, I provide examples of how `methsim` can be used to learn about DNA methylation and how it might be used to develop simulation studies to compare competing analysis methods.
\end{chapabstract}

1. Introduction
  - Why simulation?
2. Literature review
  - Simulating reads vs. $\beta$-values
  -
3. Methods
  - Implementation
4. Case study
5. Conclusions

@@

Most papers that propose a new method for downstream analysis of bisulfite-sequencing data include a simulation section. Generally, such a simulation method exists to support claims about the performance of the proposed method and is not a major feature of the paper. Consequently, the simulation model is often only briefly described and a software implementation is rarely made available. In fact, until recently, of the methods reviewed in this section, only that of \citet{Lacey:2013iy} had a software implementation available[^WGBSSuite]. As I was writing this chapter, the `WGBSSuite` software was published \citep{Rackham:2015bv}. `WGBSSuite` is the only software specifically published for the purpose of simulating whole-genome bisulfite-sequencing data when comparing methods for identifying differential methylation. `WGBSSuite` is available for download as a collection of R scripts[^WGBSSuite].

[^WGBSSuite]: Unfortunately, `WGBSSuite` is not available as an R package, which is the "fundamental unit of reproducible R code" ([http://r-pkgs.had.co.nz/](http://r-pkgs.had.co.nz/)) and would greatly simplify the installation and use of the software. Furthermore, no licence file is included in the download, meaning that it is unclear how the user is permitted to use `WGBSSuite` and whether they may modify or redistribute the code. (__TODO: Update if I received a response. I emailed the developer, Owen Rackham, asking him to add a licence. He is considering my request.__)

@@

I do not think this a useful or necessary step. In fact, I think it is counterproductive. While in truth the set of methylation loci do vary between samples due to genetic variation, it is a reasonable approximation to consider the positions of these loci as fixed. If sequence variation is required then it is easily accommodations by sampling from the set of methylation loci in the reference genome. And it is counterproductive because the aim of the simulation model is to realistically simulate methylation _levels_, not the _locations_ of these loci. Furthermore, almost all downstream analyses are _reference-based_ (see Section \ref{sec:methylated_or_unmethylated}), so it is desirable to know how these methods perform with respect to the relevant reference genome, not some simulated genome in which the location of methylation loci vary from simulation to simulation.

@@

- [x] DONE

The sequencing depth at each methylation loci, $m_{i, j} + u_{i, j}$, may be sampled from real data \citep[e.g.]{Feng:2014iq, Chen:2014jb, Dolzhenko:2014bo}, or simulated from a parametric distribution such as the Poisson \citep{Rackham:2015bv}, a rounded Gaussian distribution \citep{Xu:2013eg}, or a rounded mixture of Gamma distributions \citep{Lacey:2013iy}.

@@

- [x] DONE

The final step is to simulate the read counts, $M_{i, j}$ and $U_{i, j}$. These are based on the true underlying methylation level, $B_{i, j}$, and the sequencing depth, $m_{i, j} + u_{i, j}$. In the beta-binomial model this is done by binomial sampling where $M_{i, j} \eqd binomial(m_{i, j} + u_{i, j}, B_{i, j})$. In addition to binomial read sampling, the `WGBSSuite` software also implements (truncated) negative binomial read sampling \citep{Rackham:2015bv}, which essentially introduces overdispersion in the read counts.

@@

As demonstrated in Chapter \ref{chap:co-methylation}, the physical process of co-methylation is at the level of individual DNA fragments, not at the level of aggregate methylation levels. The observed correlations of $\beta$-values are driven by this within-fragment co-methylation. However, a simulation model of DNA methylation data need not induce correlations at the level of individual DNA fragments. Indeed, if the model does not simulate the individual fragments then it cannot simulate within-fragment co-methylation, but it still may be able to capture correlations of methylation levels by inducing dependence in the $B_{i, j}$. For example, under the Beta-Binomial model these correlations could be induced by forcing the $\mu_{i, j}$ to be spatially dependent. This idea takes its inspiration from \citet{Jaffe:2012gx} who simulate spatially correlated DNA methylation microarray data by imposing an autocorrelation structure via a lag-1 autoregressive model of the simulated $\beta$-values.

@@

@@

- [x] DONE

`methsim` is written in R and builds on the `MethylationTuples` package described in Chapter \ref{chap:wgbs_downstream_analyses}, as well as several R packages available on Bioconductor and CRAN (see the `DESCRIPTION` file for a complete list.).

@@

To simulate a 'true' methylome, the user runs the `simulate()` method on the `SimulateMethylomeParam` object. This returns a `SimulatedMethylome` object.

Once we have a 'true' methylome, we can simulate data from it. Currently, `methsim` supports the simulation of whole-genome bisulfite-sequencing reads via the `WGBSParam` class and associated `simulate()` method. Other assays, such as RRBS or microarrays may be supported in future releases. A `WGBSParam` object will contain a `SimulatedMethylome` object, along with parameters such as the read-length, sequencing coverage and error rate of the data to be simulated. When applied to a `WGBSParam` object, which includes a `SimulatedMethylome` object, the `simulate()` method can return a `SimulatedBS` object or a `MethylationTuples::MethPat` object[^namespace2]. A `SimulatedBS` object contains all simulated reads and is generally much larger than a `MethylationTuples::MethPat` object, which summarises the simulated reads for m-tuples of a particular size[^methpat]. `methsim` also includes a helper function, `asMethPat()`, to coerce a `SimulatedBS` object to a `MethylationTuples::MethPat` object.

@@

- [ ] DONE

As we have seen, DNA methylation is highly heterogeneous along the genome. Nonetheless, there are clear regions of "similarity", such as the unmethylated CpG islands and long partially methylated domains. In these locally-similar regions we might hope to model DNA methylation by a simple parametric model. `methsim` is based on the idea of segmenting the genome into 'regions of similarity', fitting a simple model to each region and then 'stitching' the results together to form the true methylome.

@@

1. Segment the methylome into regions of similarity.
2. Sample parameters for each methylation locus, $i$, based on step 1.

@@

[^pmds]: Partially methylated regions are also commonly known as partially methylated domains, but I refer to them as regions for consistency with UMRs and LMRs.

@@

`methsim` post-processes the segmentation provided by `MethylSeekR` to partition the methylome into regions of similarly, namely \emph{UMR}s, \emph{LMR}s, \emph{PMRS}s and \emph{other}s[^methylseekr_output]. Roughly speaking, _other_ regions are 'mostly methylated regions' (see Figures \ref{fig:EPISCOPE_beta_by_pm_boxplots}, \ref{fig:Lister_beta_by_pm_boxplots}, \ref{fig:Seisenberger_beta_by_pm_boxplots}, \ref{fig:Ziller_beta_by_pm_boxplots}), although \citet{Burger:2013kq} do not describe these regions as such.

@@

`methsim` uses a two-state ($1 =$ methylated or $0 =$ unmethylated), one-step Markov chain to model $\mathbf{Z} = \{ \mathbf{Z_{h}} \}_{h = 1}^{h = n_{h}}$. This allows the incorporation of within-fragment co-methylation into the simulation. The choice of a one-step process is one of computational simplicity, although I will show that this is a reasonable approximation of within-fragment co-methylation.

In Chapter \ref{chap:co-methylation} I showed that the strength of within-fragment co-methylation varies as a function of the intra-pair distance and by the genomic context. I also showed that for a pair of CpGs separated by given IPD, and in the same genomic context, that there is a large amount of variability in the strength of within-fragment co-methylation.

For these reasons I allow the transition probabilities, $\mathbf{p}$, to vary with $i$, that is, I allow the Markov chain to be spatially inhomogeneous. In particular, I allow the transition probabilities to depend on the intra-pair distance between the $i^{th}$ and $(i + 1)^{th}$ locus and on the region type, $r_{i}$ (_UMR_, _LMR_, _PMR_ or _other_), for this pair of loci[^context]. That is, $p_{i; a, b} = Pr(Z_{h, i + 1} = b | Z_{h, i} = a, IPD(i, i + 1), r_{i})$.

@@

Rather than directly modelling the transition probabilities, $\mathbf{p}$, `methsim` is parameterised $\mathbf{Z}$ by a vector of marginal probabilities, $\mathbf{B} = \{ Pr(Z_{i} = 1) \}$, and a vector of odds ratios, $\mathbf{\psi} = \{ \frac{Pr_{Z_{i + 1} = 1 | Z_{i} = 1} \times Pr_{Z_{i + 1} = 0 | Z_{i} = 0}}{Pr_{Z_{i + 1} = 1 | Z_{i} = 0} \times Pr_{Z_{i + 1} = 0 | Z_{i} = 1}} \} = \{ \frac{Pr_{Z_{i} = 1, Z_{i + 1} = 1} \times Pr_{Z_{i} = 0 , Z_{i + 1} = 0}}{Pr_{Z_{i} = 1 , Z_{i + 1} = 0} \times Pr_{Z_{i} = 0 , Z_{i + 1} = 1}} \}$. We can compute the transition probabilities, $\mathbf{p}$, from $\mathbf{B}$ and $\mathbf{\psi}$ as I now describe.

@@

Once $P_{i, i + 1}$ is computed, the desired transition probability is obtained by dividing the appropriate element of $P_{i, i + 1}$ by the appropriate marginal probability. To continue our example, suppose $Z_{i} = 0$, then the probability that $Z_{i + 1}$ is also to zero is given by $Pr(Z_{i + 1} = 0 | Z_{i} = 0) = \frac{Pr(Z_{i = 0}, Z_{i + 1} = 0)}{Pr(Z_{i} = 0)} = \frac{0.155}{0.3} = 0.517$ to three decimal places. In fact, `methsim` only stores $Pr(Z_{i + 1} = 1 | Z_{i} = 0)$ and $Pr(Z_{i + 1} = 1 | Z_{i} = 1)$ since Pr(Z_{i + 1} = 0 | Z_{i} = 0) = 1 - Pr(Z_{i + 1} = 1 | Z_{i} = 0)$ and $Pr(Z_{i + 1} = 0 | Z_{i} = 1) = 1 - Pr(Z_{i + 1} = 1 | Z_{i} = 1)$.

To simulate values of $\mathbf{B}$ and $\mathbf{\psi}$, `methsim` uses the empirical distributions of their estimators, $\mathbf{\beta}$ and $\mathbf{\widehat{\psi}}$, in the input methylome. More specifically, $\mathbf{B}$ is based on $\beta_{i | r_{i}}$ and $\mathbf{\psi}$ is based on $\widehat{\psi_{IPD(i, i+1) | r_{i}}}$, the empirical distributions of $\beta_{i}$ and $\psi_{IPD(i, i + 1)}$ conditional on the region type of the $i^{th}$ locus, $r_{i}$, and the intra-pair distance of the the $i^{th}$ and $(i + 1)^{th}$ locus, $IPD(i, i+1)$, respectively. These empirical distributions are shown for the \emph{ADS} sample in Figures \ref{fig:ADS_beta_by_pm_boxplots} and __FIGURE__.

@@

__TODO: Figure of ADS within-fragment co-methylation__

These same distributions for all the _EPISCOPE_, _Lister_, _Seisenberger_ and _Ziller_ samples are shown in __FIGURES__.

There are several ways to simulate values of $\mathbf{B}$ and $\mathbf{\psi}$ based on these empirical distributions. The simplest is to sample each $B_{i}$ and $\psi_{i}$ directly from these empirical distributions. Alternatively, we could take a more model-based approach. For example, we could specify a  parametric model of the $B_{i}$ where all $i$ in a region have the same mean and variance (the mean and variance being based on the empirical distribution of $\beta$-values) or where the $B_{i}$ in each region come from an autoregressive process.

The simulation of $\mathbf{\psi}$ requires further explanation. The empirical distributions of $\mathbf{\widehat{\psi}}$ will only include values for small $IPD$s since these are estimated from individual sequencing reads (see Chapter \ref{chap:co-methylation} for details). When simulating a methylome, we will encounter 2-tuples with much larger $IPD$s and the question is how to deal with these? Again, there are several options. We might assume a parametric form for the distribution of $\mathbf{\psi}$ as a function of $IPD$, such as a exponential decay towards $\psi = 1$ (corresponding to independence).

I explore different methods for simulating $\mathbf{B}$ and $\mathbf{\psi}$ in Section \ref{sec:results}.

@@

`methsim` uses a simple Poisson-based method for simulating bisulfite-sequencing. The user specifies the desired read length, the average sequencing coverage[^read_length], the error rate ($\epsilon$, which includes both sequencing error and bisulfite-conversion error), and provides the simulated 'true' methylome.

@@

\subsection{Construct the output}

@@

To simulate these type of experiments, we need to be able to control the variation between the `SimulatedWGBS` objects. In this example, we probably want the three samples in each group to be more similar to each other than to the three samples in the other group. Essentially, we need to be able to control the within-group and between-group variation.

@@

We might also consider more subtle ways of introducing variation. For example, we might use the same `MethylomeParam` object but vary how we simulate $\mathbf{B}$ and $\mathbf{\psi}$ between calls to `simulate()`.

and the same average methylation level for each region but vary the strength of co-methylation. This could be achieved by altering the $LOR_{(i, i+1) | r_{i}, IPD(i, i+1)}$ distributions between two `SimulateMethylomeParam` objects while holding the rest constant. Or we might vary the weights, $W$, to simulate more or less heterogeneous samples.

In Section ref{sec:results} I explore how much variation is created using these different techniques.

@@

__UP TO HERE: See what below can be salvaged__

@@

Shown below are the running times and memory usage involved in simulating a $30\times$ whole-genome bisulfite-sequencing sample using the _ADS_ `MethylomeParam` object included in `methsim`. Using up to 8 CPU cores in parallel, it takes less than 5 minutes to simulate the underlying 'true' methylome and a further __HOW LONG__ to simulate a single $30 \times$ sequencing coverage whole-genome bisulfite-sequencing assay of this methylome[^parallel].

[^parallel]: NB the creation of the `SimulatedMethylome` via `simulate(ADS, BPPARAM)` currently only uses a single CPU.

__UP TO HERE: simulate-WGBSParam-method seems to be causing a memory leek and resulting crash. Running valgrind in reduced problem to investigate.__

```{r, echo = TRUE, eval = FALSE}
# Load necessary packages
suppressPackageStartupMessages(library(methsim))
suppressPackageStartupMessages(library(BiocParallel))
suppressPackageStartupMessages(library(pryr))

# Use up to 8 cores in parallel computations.
BPPARAM <- MulticoreParam(workers = 8L)

# Set seed for reproducibility.
set.seed(666)

# ADS: the MethylomeParam object.
pryr::object_size(ADS)
#> 16.5 MB

# sm: the SimulatedMethylome object
system.time(l_sm <- simulate(ADS, BPPARAM = BPPARAM))
#> Currently only simulates CpG methylation.
#> Currently only simulates unstranded methylomes.
#> Simulating 1 methylome..
#>    user  system elapsed
#> 186.680   9.676 270.780
sm <- l_sm[[1]]
pryr::object_size(sm)
#> 676 MB

# wgbsp: the WGBSParam object.
# NOTE: Basically all the memory used by the WGBSParam object is the
# SimulatedMethylome object.
wgbsp <- WGBSParam(SimulatedMethylome = sm,
                   AveCov = 30,
                   ErrorRate = 0.01,
                   SequencingType = 'SE',
                   ReadLength = 100)
pryr::object_size(wgbsp)
#> 676 MB

# sbs: the SimulatedBS object.
system.time(sbs <- simulate(wgbsp, BPPARAM = BPPARAM))
#> Simulating 1 bisulfite-sequencing sample...

#>     user   system  elapsed
#> 2716.503  174.591  548.122
pryr::object_size(sbs)
#> 10.1 GB

# methpat: the MethylationTuples::MethPat object of one-tuples.
# NOTE: information is lost when a SimulatedBS object is coerced to a MethPat
# object.
system.time(methpat <- asMethPat(sbs,
                                 sampleName = "sim_ADS",
                                 size = 1L,
                                 BPPARAM = BPPARAM))
#>    user  system elapsed
#> 557.900  34.779  48.944
pryr::object_size(methpat)
#> 451 MB
```

These results are indicative of simulating high-coverage whole-genome bisulfite-sequencing for human-sized genomes. Simulations using a smaller genome or lower sequencing coverage will have faster running times and lower memory usage.

@@

- [ ] DONE

A central pillar of `methsim` is that many of the simulation parameters are sampled from real data. At a bare minimum, the first test of `methsim` is that simulated data are similar to the real data on which they are based. To assess this, I compare several summary measures between the simulated data and the real data. The simplest of these are summaries are the distributions of $\beta$-values and within-fragment co-methylation, which are explicitly sampled in `methsim`. In contrast, the spatial correlation of $\beta$-values is not explicitly modelled in `methsim`; rather, we hope that `methsim` recapitulates this phenomenon by appropriate modelling of within-fragment co-methylation and choice of weights, $\mathbf{W}$.

Once it is established that `methsim` passes this minimal test, I use `methsim` to explore models of within-fragment co-methylation, the choice of the weights, $\mathbf{W}$, and their respective effects on the summary measures in order to identify areas of improvement in `methsim`. Finally, I demonstrate how `methsim` might be used to simulate data for use in a study comparing methods for identifying differential methylation.

\subsection{Summary measures of simulated data}

- [ ] DONE

\subsection{Models of within-fragment co-methylation}

- [ ] DONE

- Apparent co-methylation arises because of dependence of $B_{i}$, even when $Z_{i}$ are independent! E.g., set within-fragment co-methylation to zero but have $B_{i}$ fixed for a region results in independent $Z_{i}$ but dependent $B_{i}$.
  - What to make of this?
  - What if we sample $B_{i}$ from $\beta_{i | r_{i}}$ instead of using a fixed value for all loci in a region?

\subsection{The choice of $\mathbf{W}$}

\subsection{Simulating differential methylation}

- [ ] DONE

\section{Conclusions}

- [ ] DONE

\subsubsection{Comparison to other methods}

__TODO: Work the comparison to other methods into my conclusions__

\citet{Lacey:2013iy} supply R scripts to implement their simulation method ([http://rrbs-sim.r-forge.r-project.org/](http://rrbs-sim.r-forge.r-project.org/)). __TODO: Try these scripts, see how long they take__. The running times reported in the original publication are shown in __TABLE__ for a single simulation. __TODO: To simulate a single RRBS dataset would take HOW LONG__. The simulation scripts could be made more user-friendly by creating an R package, rather than supplying the software as a collection of scripts, and would benefit from some documentation.

\begin{table}[h]
\centering
\begin{tabular}{@{}lll@{}}
\toprule
Number of samples & Number of sites & Running time (seconds) \\ \midrule
8                 & 1000            & \textless 1            \\
8                 & 5000            & 13                     \\
8                 & 10000           &                        \\ \bottomrule
\end{tabular}
\caption{Reported running times for simulation methodology of \citet{Lacey:2013iy}. Simulations were run on "on a desktop with a 3.4 GHz processor and 10 GB RAM running 64-bit Windows 7".}
\label{tab:Lacey_runtime}
\end{table}

__TODO: Compare `methsim` to `WGBSSuite`.

# TODO:

- [ ] Decide on a variable for sequencing depth; $m_{i, j} + u_{i, j}$ is kinda confusing. Perhaps $d_{i, j} / D_{i, j}$?

>>>>

\begin{chapabstract}
This chapter describes the `methsim` software to simulate DNA methylation data. `methsim` incorporates a model of within-fragment co-methylation and simulates individual sequencing reads, which sets it apart from existing software. We explore different parameter choices and highlight promising directions of research, such as using `methsim` to simulate data in order to compare methods for identifying differential methylation.
\end{chapabstract}

@@

Most papers that propose a new method for downstream analysis of bisulfite-sequencing data include a simulation section. Generally, such a simulation method exists to support claims about the performance of the proposed method and is not a major feature of the paper. Consequently, the simulation model is often only briefly described and a software implementation is rarely made available. In fact, until recently, of the methods reviewed in this section, only that of \citet{Lacey:2013iy} had a software implementation available. As I was writing this chapter, the `WGBSSuite` software was published \citep{Rackham:2015bv}. `WGBSSuite` is the only software specifically published for the purpose of simulating whole-genome bisulfite-sequencing data when comparing methods for identifying differential methylation. `WGBSSuite` is available for download as a collection of R scripts[^WGBSSuite].

[^WGBSSuite]: Unfortunately, `WGBSSuite` is not available as an R package, which is the "fundamental unit of reproducible R code" ([http://r-pkgs.had.co.nz/](http://r-pkgs.had.co.nz/)) and would greatly simplify the installation and use of the software. Furthermore, no license file is included in the download, meaning that it is unclear how the user is permitted to use `WGBSSuite` and whether they may modify or redistribute the code.

@@

I do not think this a useful or necessary step and may even be counterproductive. While in truth the set of methylation loci do vary between samples due to genetic variation, it is a reasonable approximation to consider the positions of these loci as fixed. If sequence variation is required then it is easily accommodations by sampling from the set of methylation loci in the reference genome. Furthermore, the aim of the simulation model is to realistically simulate methylation _levels_, not the _locations_ of these loci.

Almost all downstream analyses are _reference-based_ (see Section \ref{sec:methylated_or_unmethylated}), so it is desirable to know how these methods perform with respect to the relevant reference genome, not some simulated genome in which the location of methylation loci vary from simulation to simulation.

@@

The sequencing depth at each methylation loci, $d_{i, j}$, may be sampled from real data \citep[e.g.]{Feng:2014iq, Chen:2014jb, Dolzhenko:2014bo}, or simulated from a parametric distribution such as the Poisson \citep{Rackham:2015bv}, a rounded Gaussian distribution \citep{Xu:2013eg}, or a rounded mixture of Gamma distributions \citep{Lacey:2013iy}.

@@

The final step is to simulate the read counts, $M_{i, j}$ and $U_{i, j}$. These are based on the true underlying methylation level, $B_{i, j}$, and the sequencing depth, $d_{i, j}$. In the beta-binomial model, this is done by binomial sampling where $M_{i, j} \eqd binomial(m_{i, j} + u_{i, j}, B_{i, j})$. In addition to binomial read sampling, the `WGBSSuite` software also implements (truncated) negative binomial read sampling \citep{Rackham:2015bv}, which essentially introduces overdispersion in the read counts.

@@

Most simulation methods do not simulate individual reads and so cannot simulate within-fragment co-methylation. Instead, they capture correlations of methylation levels by inducing dependence in the $B_{i, j}$. For example, under the Beta-Binomial model these correlations could be induced by forcing the $\mu_{i, j}$ to be spatially dependent. This idea takes its inspiration from \citet{Jaffe:2012gx} who simulate spatially correlated DNA methylation microarray data by imposing an autocorrelation structure via a lag-1 autoregressive model of the simulated $\beta$-values.

@@

In what follows we will simulate data based on the _ADS_ methylome from the _Lister_ dataset. We focus on simulating autosomal data, since the sex chromosomes and mitochondrial DNA have very different methylation dynamics.

@@

`methsim` is written in R and builds on the `MethylationTuples` package described in Chapter \ref{chap:wgbs_downstream_analyses}, as well as several R packages available on Bioconductor and CRAN (see the `DESCRIPTION` file for a complete list.). It is currently a very experimental package and is therefore not yet published on Biocondutor, but its development can be followed at [https://github.com/PeteHaitch/methsim](https://github.com/PeteHaitch/methsim).

@@

To simulate a true methylome, the user runs the `simulate()` method on the `MethylomeParam` object. This returns a `SimulatedMethylome` object.

Once we have a true methylome, we can simulate data from it. `methsim` currently supports the simulation of whole-genome bisulfite-sequencing reads via the `WGBSParam` class and associated `simulate()` method. Other assays, such as RRBS or microarrays could in principle be supported. A `WGBSParam` object will contain a `SimulatedMethylome` object, along with parameters such as the read-length, sequencing coverage and error rate of the data to be simulated. When applied to a `WGBSParam` object, the `simulate()` method returns a `SimulatedBS` object or a `MethylationTuples::MethPat` object[^namespace2]. A `SimulatedBS` object contains all simulated reads and is generally a large object (on the order of $10$ GB). By contrast, the `MethylationTuples::MethPat` object, which summarises the simulated reads for m-tuples of a particular size[^methpat], is much smaller (on the order of $100$ MB) but does not contain the full information of the simulation since read-level data are lost. `methsim` also includes a helper function, `asMethPat()`, to coerce a `SimulatedBS` object to a `MethylationTuples::MethPat` object.

@@

As we have seen, DNA methylation is highly heterogeneous along the genome. Nonetheless, there are clear regions of "similarity", such as the unmethylated CpG islands and long partially methylated domains. In these locally-similar regions we might hope to model DNA methylation by a simple parametric model. `methsim`, like other simulation methods, is based on the idea of segmenting the genome into 'regions of similarity', fitting a simple model to each region and then 'stitching' the results together to form the true methylome.

@@

1. Segment the methylome into regions of similarity.
2. Sample parameters for each methylation locus, $i$, based on the segmentation.

@@

[^pmds]: Partially methylated regions are also commonly known as partially methylated domains, but we will refer to them as 'regions' for consistency with UMRs and LMRs.

@@

`methsim` post-processes the segmentation provided by `MethylSeekR` to partition the methylome into regions of similarity, namely \emph{UMR}s, \emph{LMR}s, \emph{PMRS}s and \emph{other}s[^methylseekr_output]. Roughly speaking, _other_ regions are 'mostly methylated regions' (see Figures \ref{fig:EPISCOPE_beta_by_pm_boxplots}, \ref{fig:Lister_beta_by_pm_boxplots}, \ref{fig:Seisenberger_beta_by_pm_boxplots}, \ref{fig:Ziller_beta_by_pm_boxplots}), although \citet{Burger:2013kq} do not describe these regions as such.

@@

`methsim` uses a two-state ($1 =$ methylated or $0 =$ unmethylated), one-step Markov chain to model $\mathbf{Z} = \{ \mathbf{Z_{h}} \}_{h = 1}^{h = n_{h}}$. This allows the incorporation of within-fragment co-methylation into the simulation. The choice of a one-step process is one of computational simplicity and is a reasonable approximation for most of the genom given the available data (see Chapter \ref{chap:co-methylation}).

In Chapter \ref{chap:co-methylation} we saw that the strength of within-fragment co-methylation varies as a function of the intra-pair distance and by the genomic context. For this reason I allow the transition probabilities, $\mathbf{p}$, to vary with $i$, that is, I allow the Markov chain to be spatially inhomogeneous. In particular, I allow the transition probabilities to depend on the intra-pair distance between the $i^{th}$ and $(i + 1)^{th}$ locus and on the region type, $r_{i}$ (_UMR_, _LMR_, _PMR_ or _other_), for this pair of loci[^context]. That is, $p_{i; a, b} = Pr(Z_{h, i + 1} = b | Z_{h, i} = a, IPD(i, i + 1), r_{i})$.

@@

Rather than directly modelling the transition probabilities, $\mathbf{p}$, `methsim` is parameterised by a vector of marginal probabilities, $\mathbf{B} = \{ Pr(Z_{i} = 1) \}$, and a vector of odds ratios, $\mathbf{\psi} = \{ \frac{Pr_{Z_{i + 1} = 1 | Z_{i} = 1} \times Pr_{Z_{i + 1} = 0 | Z_{i} = 0}}{Pr_{Z_{i + 1} = 1 | Z_{i} = 0} \times Pr_{Z_{i + 1} = 0 | Z_{i} = 1}} \} = \{ \frac{Pr_{Z_{i} = 1, Z_{i + 1} = 1} \times Pr_{Z_{i} = 0 , Z_{i + 1} = 0}}{Pr_{Z_{i} = 1 , Z_{i + 1} = 0} \times Pr_{Z_{i} = 0 , Z_{i + 1} = 1}} \}$. We can compute the transition probabilities, $\mathbf{p}$, from $\mathbf{B}$ and $\mathbf{\psi}$.

@@

Once $P_{i, i + 1}$ is computed, the desired transition probability is obtained by dividing the appropriate element of $P_{i, i + 1}$ by the appropriate marginal probability. To continue our example, suppose $Z_{i} = 0$, then the probability that $Z_{i + 1}$ is also zero is given by $Pr(Z_{i + 1} = 0 | Z_{i} = 0) = \frac{Pr(Z_{i = 0}, Z_{i + 1} = 0)}{Pr(Z_{i} = 0)} = \frac{0.155}{0.3} = 0.517$ (to three decimal places). In fact, `methsim` only stores $Pr(Z_{i + 1} = 1 | Z_{i} = 0)$ and $Pr(Z_{i + 1} = 1 | Z_{i} = 1)$ since Pr(Z_{i + 1} = 0 | Z_{i} = 0) = 1 - Pr(Z_{i + 1} = 1 | Z_{i} = 0)$ and $Pr(Z_{i + 1} = 0 | Z_{i} = 1) = 1 - Pr(Z_{i + 1} = 1 | Z_{i} = 1)$.

To simulate values of $\mathbf{B}$ and $\mathbf{\psi}$, `methsim` uses the empirical distributions of their estimates, $\mathbf{\beta}$ and $\mathbf{\widehat{\psi}}$, in the input methylome. More specifically, $\mathbf{B}$ is based on $\beta_{i | r_{i}}$ (the empirical distribution of $\beta_{i}$ conditional on the region type of the $i^{th}$ locus, $r_{i}$) and $\mathbf{\psi}$ is based on $\widehat{\psi_{MH}}$. We have already seen many examples of the distribution of $\widehat{\psi_{MH}}$ in Chapter \ref{chap:co-methylation}. The empirical distributions $\beta_{i | r_{i}}$ for the _ADS_ sample are shown in Figure \ref{fig:ADS_beta_by_pm_boxplots}.

@@

There are many ways to simulate values of $\mathbf{B}$ and $\mathbf{\theta} = \mathbf{log{\psi}}$ based on these empirical distributions. I have so far only had time to explore a few simple methods and this remains ongoing work. We will discuss the results from these five models (m1, m2, m3, m4, m5):

\begin{description}
  \item[m1] All loci in each region have the same $B$, which is sampled from $\beta_{i | r_{i}}$. $\theta$ sampled from $Gaussian(\mu_{IPD}, \sigma_{IPD}^{2})$, where $\mu_{IPD}$ and $\sigma_{IPD}$ are plug-in estimates computed from the distribution of chromosome-level $\widehat{\theta_{MH}}$ (autosomes only).
  \item[m2] All loci in each region have the same \textbf{average} $B$, which is sampled from $\beta_{i | r_{i}}$; this is then perturbed by a $Gaussian(0, \sigma_{B}^2)$ random variable. This is similar to what is done by `WGBSsuite`. Here we use $\sigma_{B}^{2} = B \times (1 - B)$. The resulting $\mathbf{B}$ is truncated so that all values lie between 0.01 and 0.99 (necessary to avoid issues in the iterative proportional fitting algorithm at the boundaries). $\theta$ simulated as in m1.
  %\item[m3] $\mathbf{B}$ simulated as in m2 but with the perturbations in each region simulated from a first-order autoregressive process with coefficient equal to 0.5. $\theta$ simulated as in m1.
  %\item[m4] As for m3 but where the sample is a 50:50 mixture, i.e., two independent simulations of $\mathbf{B}$ and $\mathbf{\theta}$ that are sampled from with equal probability.
  %\item[m5] As for m3 but with $\mathbf{theta} = 0$ to simulate independence of methylation states within fragments.
\end{description}

The simulation of $\mathbf{\theta} = \mathbf{log{\psi}}$ requires further explanation. The empirical distributions of $\mathbf{\widehat{\theta}}$ will only include values for small $IPD$s since these are estimated from individual sequencing reads (see Chapter \ref{chap:co-methylation} for details). When simulating a methylome, we will encounter 2-tuples with much larger $IPD$s and the question is how to deal with these?

Again, there are several options. For example, we might assume a parametric form for the distribution of $\mathbf{\psi}$ as a function of $IPD$, such as a exponential decay towards $\theta = 0$ (corresponding to independence). Given the quality and read-lengths of the _ADS_ data, I have elected to use the empirical $\widehat{\theta_{MH}}$ for pairs with $IPD \leq 180$ and set $\mu_{IPD} = 0$ for those with $IPD > 180$[^sigma]. While obviously a gross simplification, the vast majority of $NIL = 0$ CpG pairs have an $IPD \leq 180$ and so will be unaffected by this 'independence' assumption.

[^sigma]: $\sigma_{IPD}$ is given by the median absolute deviation of $\sigma_{IPD}$ for $IPD \leq 180$.

There is an increase in complexity for models m1 through to m4, while m5 is there as a sanity check that the within-fragment co-methylation model is working as intended. The limitations of these models may already be clear, but we will discuss these further in Section \ref{sec:methsim_results}.

@@

`methsim` uses a simple Poisson-based method for simulating bisulfite-sequencing. The user specifies the desired read length, the average sequencing coverage[^read_length], the error rate ($\epsilon$, which includes both sequencing error and bisulfite-conversion error), and provides the simulated 'true' methylome. We will use $200$ bp, single-end reads with an average coverage of $30 \times$ in what follows.

@@

\subsection{Constructing the output}

@@

To simulate these type of experiments, we need to be able to control the variation between the resulting `SimulatedWGBS` objects. In this example, we probably want the three samples in each group to be more similar to each other than to the three samples in the other group. Essentially, we need to be able to control the within-group and between-group variation.

@@

We might also consider more subtle ways of introducing variation. For example, we might use the same `MethylomeParam` object but vary how we simulate $\mathbf{B}$ and $\mathbf{\psi}$ between calls to `simulate()`. This avenue remains to be explored.

@@

@@

Using up to 8 CPU cores in parallel, it takes less than 9 minutes to simulate a single $30 \times$ sequencing coverage whole-genome bisulfite-sequencing assay from a 'true' methylome. These results are indicative of simulating high-coverage whole-genome bisulfite-sequencing for human-sized genomes. Simulations using a smaller genome or lower sequencing coverage will have faster running times and lower memory usage.

@@

A central feature of `methsim` is that the simulation parameters are sampled from an input sample. At a bare minimum, the first test of `methsim` is that simulated data are similar to the real data on which they are based. To assess this, we compare several summary measures between the simulated data and the real data.

__TODO: Add in the _ADS_ data to the simulated data__

The simplest of these are summaries of the distributions of $\beta$-values and within-fragment co-methylation, which are explicitly sampled in `methsim`. We see that all of m1, m2, m3, m4 and m5 sufficiently capture the bimodality of $\beta$-values (__FIGUREs__) and their relationship to CpG islands (__FIGUREs__).

__FIGURE__ plots the estimates of within-fragment co-methylation. Those models that include within-fragment co-methylation (m1, m2, m3 and m4) bear close resemblance to the real _ADS_ data, although are perhaps to homogeneous across chromosomes. Reassuringly, we see no evidence of within-fragment co-methylation in model 5, which simulates independence of within-fragment methylation states.

In contrast, the spatial correlation of $\beta$-values is not explicitly modelled in any of models m1 to m5. While m3, m4 and m5 do include some spatial dependence of the $\mathbf{B}$, these are independent of $IPD$ and are intended mostly to analyse the effect of the independence condition in m2. __TODO: Describe these crap results.__

\section{Summary}

Simulating DNA methylation data is complicated by its heterogeneity. This heterogeneity exists is in multiple 'directions': along the genome, between cells in a sample, and between subjects. As more data become available, we can get a better handle on the causes of the variation but, for now, it remains challenging.

`methsim` provides a framework that I have used to experiment with models for simulating DNA methylation data. The models I have currently explored capture some key aspects of the data including, for the first time, within-fragment co-methylation. Unfortunately, while these results are promising on some fronts, there is clearly much work to be done on others.

Most notably, and frustratingly, I do not yet have an adequate model for the correlations of the true methylation levels, $\mathbf{B}$. The model of \citet{Lacey:2013iy} is promising, although may be computationally infeasible for whole-genome data given the timings reported in the original publication. The model used by `WGBSsuite` \citet{Rackham:2015bv} is similar to m2, so I suspect it will produce similar results, but this requires further investigation. I continue to explore how alternative models of $\mathbf{B}$ might be integrated into `methsim` to improve this aspect of the simulation.
